Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning

### 2017-02-14 paper.tex
Cut it down to 12 pages.
Please @mariosky and @pacastillo have a look. @pacastillo, fork and to a pull request, please.

The structure of this work is organized as follows: next Section 
presents a series of works related to the proposed method in this paper; Section
\ref{sec:method} describes the proposed method for the recognition of affective
states in a web based learning environment for the teaching of programming
languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
proposed method, and results are discussed in Section \ref{sec:res}  % There is no results section so far
Finally, we draw some conclusions about our methodology and
experiments.

@@

Elliott, Rickel y Lester \cite{elliott1999lifelike,d2008autotutor} propose the integration of an affective reasoner
with an agent in a virtual environment. In this work an agent called Steve
responds emotionally to other agents and interactive users. The agent simulates
his emotions through different multimedia modes including facial expressions and
speech. Affective reasoning agents were used for training, by putting students
in work related situations. Agents could react to the behavior of students, for
instance if a student was being careless in a task and was in a dangerous
situation, Steve would show distress or fear. Also in a virtual environment the
work of McQuiggan, Robison \& Lester \cite{mcquiggan2010affective} extends this line of research by
investigating the affective transitions that occur throughout narrative-centered
learning experiences. The analysis of affective state transitions in this work
replicated the findings by D’Mello et al. \cite{d2008autotutor} and Baker et al. 
\cite{rodrigo2009affective} where
also engagement/flow dominated self-reported affect. For outdoor environments
Shen, Wang \& Shen \cite{shen2009affective} augment a pervasive e-learning platform with affective
recognition.  The results about emotion recognition from physiological signals
achieved a best-case accuracy (86.3\%) for four types of learning emotions.

Bosch, D'Mello \& Mills \cite{bosch2013emotions} analyzed the relationship between affective states
and performance of novice programmers when they were learning the basics of
computer programming in the Python language. The results of their study
indicated that the more common emotions students experienced were engaged
(23\%), confusion (22\%), frustration (14\%), and boredom (12\%). It was useful
to consider these results, as it presented evidence of what affective states
need to be targeted in order to obtain less biased data. For example, if a less
common emotion was chosen, a classifier could opt to classify any feature vector
as not experiencing such emotion. Nevertheless, the classifier would obtain
accurate results, although the classifier would be inaccurate at determining if
a feature vector was actually experiencing the given affective state. Similar to
the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
behaviors relate to student's achievement within a basic computer science
course. The authors found that confusion, boredom and engagement in IDE-related
(on-task) conversation are associated with lower achievement.

There has been very little research reported on the effectiveness of the use of
keyboard and mouse dynamics as a sensory channel for affective recognition, and
the few have not been focused on programming. The preliminary work by Zimmermann
et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
mouse) with an emotional state. Several physiological parameters were measured
concurrent with the task. The parameters included respiration, pulse, skin
conductance level, and corrugation activity. The task assigned to subjects was to
shop on an e-commerce website for office-supplies and finally write a predefined
message to the shop operator. Subjects where watching videos in order to change
their emotional state during the experiment. The method was validated using
self-reported emotions using the self-assessment-manikin (SAM), devised by Lang
\cite{lang1980behavioral}, designed to assess the dimensions valence, arousal and dominance
directly by means of three sets of graphical manikins. Results show that the
film clips were effective in inducing the expected mood changes, but no further
empirical results are presented. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
uses sensory data based on the time elapsed between each key press, the task
given to users was to write a free-text and used linguistic features in order to
recognize both physical and emotional stress. The results show a classification
accuracy of 62.5\% for physical stress and 75\% from emotional stress; authors
argue that these results are comparable with other approaches in affective
computing. They also stress that their methods must be validated further in
other contexts. Moods may have an impact on programmer’s performance according
to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
on the basis of information regarding the programmer’s use of the keyboard and
mouse, and to integrate them into development environments that can improve
programmer performance. They briefly describe a further experiment that could
use keyboard and mouse but only as future work. There are other studies about
the behavior of programmers, which are not directly concerned with affective
states but are nevertheless important to their performance. Eye tracking in
computing education is proposed in the work Busjahn et al., \cite{busjahn2014eye}, and it is
also used for assessing learner´s comprehension of C++ and Python programs in
Turner et al., \cite{turner2014eye}. Blikstein \cite{blikstein2011using} proposed 
an automated technique to
assess, analyze and visualize students learning computer programming. Blikstein
employs different quantitative techniques to extract students’ behaviors and
categorize them in terms of programming experience.

% \begin{figure}[!t] 
% \centering 
% \includegraphics[width=2.5in]{KMDAffective.png} 
% \caption{Code Execution and Affect Recognition Architecture.}
% \label{fig_process} 
% \end{figure}

@@

Although the use of Keystroke Dynamics (KD) is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed text, their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models.

Regarding Mouse Dynamics (MD), some research has been conducted for the
identification of affective states, although, as with the case of KD, MD is
mainly used as a biometric measure for authentication. Salmeron-Majadas, Santos
and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
five different classification algorithms. Bakhtiyari y Husain \cite{bakhtiyari2014fuzzy} discuss a
method based on fuzzy models for the recognition of emotions through KD, MD and
touch-screen interactions. For a broad review of emotion recognition methods
based on KD and MD, the work by Kolakowska \cite{kolakowska2013review} is recommended.

@@

To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner's affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded how many attempts
a student required before successfully solving an exercise. This number of
attempts is also included in the feature vector. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying} and are shown in Table \ref{tab_features}. (The code used for this step is available at:
https://------.---/-------/-----------------------). %anonymize

Once feature vectors are obtained from an experiment, the generated dataset is
normally used as training data for a classifier. Researchers of affective
recognition have used a wide variety of classification algorithms. As a proof of
concept four well known classification algorithms where compared: k-Nearest
Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
propagation, a na\"ive Bayes classifier and finally a decision trees algorithm for
rational data (J-48). Details for these algorithms, can be found in the textbook
by Tan, Steinbach, \& Kumar \cite{tan2006introduction}. The accuracies obtained by these algorithms
as well as the parameters used are reported in the next section.

\section{Experiment}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system. The context of use is a learning environment for programmers. In
particular for learning activities consisting in writing short programs
interactively. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many exercises as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the exercises or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure}[!t] 
\centering 
\includegraphics[width=2in]{Completed.png} 
\includegraphics[width=2in]{classDist.png} 
\caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}
\label{fig_completed} 
\end{figure}

@@

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network, but the fact that they needed Facebook kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services; however, the use of Facebook gave researchers access to public
information about the subjects. Their ages were in the range of 18 to
30 years. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, participants had
different levels of experience. Results from
the initial survey is shown in Table \ref{tab_results}. %Please use reference. Fig 4
                                %is completed activities. I think this
                                %should be Fig 5. 
\begin{table*}[!t]
\centering
\caption{ Survey question: How many years of programming experience do you have? }
\label{tab_results}
    \begin{tabular}{ | c | l | }
    \hline
    Years Programming          & Number of Learners \\
    \hline
   		0   &  8 \\
    \hline
    	1   &  13\\
    \hline
    	2-4  & 15\\
    \hline
    	5+   & 22\\
    \hline
    \end{tabular}

\end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy (kappa) }
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% 0.50 & 80.48\% 0.51 & 76.14\% 0.43 & 78.43\% 0.45 \\
    \hline
    Relaxation      & 71.10\% 0.40 & 72.57\% 0.45 & 69.14\% 0.37 & 71.24\% 0.34\\
    \hline
    Distraction     & 70.33\% 0.43 & 71.00\% 0.43 & 74.00\% 0.50 & 72.52\% 0.39\\
    \hline
    Frustration     & 74.71\% 0.49 & 73.86\% 0.45 & 72.62\% 0.42 & 78.00\% 0.50\\
    \hline
    Boredom         & 74.71\% 0.47 & 84.57\% 0.59 & 83.81\% 0.58 & 76.19\% 0.45\\
    \hline
    \end{tabular}
\end{table*}

@@

\section{Results}
\label{sec:res} 
In Table \ref{tab_performance} the performance of each of the four classifiers is together 
with the kappa coefficients. A 10-Fold cross validation was used. 
The accuracies and kappa coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}. %A reference!!! - JJ  
												 %Added	
In a fixed-text method, subjects are asked to write specific texts.											 
The results obtained with this method are satisfactory, considering that learners
were writing a program, a task comparable with free-text writing, where there
is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
In the case of the kappa coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of kappa were close to 0.5, is important to consider the values of kappa
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted kappa
statistic for the kind of problem. The artificial neural network % Which one? - JJ
give the highest
accuracy in the classification of frustration.

\section{Conclusion}
The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use would be learning environment for programmers. In particular for learning
activities consisting in writing short programs interactively. An experimental
approach was adopted with this aim. The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and kappa below their counterparts of fixed-text, and in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the feature vectors. 
While these are promising results further experiments are needed. Other experiments
should focus on novice programmers.
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective?

>>>>

The structure of this work is organized as follows: next Section 
presents a series of works related to the proposed method in this paper; Section
\ref{sec:method} describes the proposed method for the recognition of affective
states in a web based learning environment for the teaching of programming
languages; next in Section \ref{sec:exp} we explain the experimental evaluation of the
proposed method and results.
Finally, we draw some conclusions about our methodology and
experiments.

@@

Bosch, D'Mello \& Mills \cite{bosch2013emotions} analyzed the relationship between affective states
and performance of novice programmers learning the basic Python. The results of their study
indicated that the most common emotions students experienced were engaged
(23\%), confusion (22\%), frustration (14\%), and boredom (12\%). It was useful
to consider these results, as it presented evidence of what affective states
need to be targeted in order to obtain less biased data. Similar to
the previous work, Rodrigo et al., \cite{rodrigo2009affective} observed which affective states and
behaviors relate to student's achievement within a basic computer science
course. The authors found that confusion, boredom and engagement in IDE-related
(on-task) conversation are associated with lower achievement. Measuring mood is important, since it may have an impact on programmer’s performance according
to Khan, Hierons y Brinkman \cite{khan2007mood}. It may be possible to detect moods
on the basis of information regarding the programmer’s use of the keyboard and
mouse, and to integrate them into development environments that can improve
programmer performance.

There has been very little research reported on the effectiveness of the use of
keyboard and mouse dynamics as a sensory channel for affective recognition, and
the few have not been focused on programming. The preliminary work by Zimmermann
et al. \cite{zimmermann2003affective} describes a method to correlate user’s interactions (keyboard and
mouse) with an emotional state, meauring different physiological parameters. The work of Vizer, Zhou \& Sears \cite{vizer2009automated} also
uses sensory data based on the time elapsed between each key press, the task
given to users was to write a free-text and used linguistic features in order to
recognize both physical and emotional stress. The results show a classification
accuracy of 62.5\% for physical stress and 75\% from emotional stress; authors
argue that these results are comparable with other approaches in affective
computing.

@@

Although the use of Keystroke Dynamics (KD) is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed text, their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models. A similar situation exists in the case of Mouse Dynamics (MD), which is used mainly for authentication; however, Salmeron-Majadas, Santos
and Boticario \cite{salmeron2014exploring} use both MD and KD to predict four affective states using
five different classification algorithms. Bakhtiyari y Husain \cite{bakhtiyari2014fuzzy} discuss a
method based on fuzzy models for the recognition of emotions through KD, MD and
touch-screen interactions. For a broad review of emotion recognition methods
based on KD and MD, the work by Kolakowska \cite{kolakowska2013review} is recommended.

@@

To calculate the average and standard deviations of these key presses, the delays
between a key-down and a key-up event of the left button clicks are used. In
addition to these averages and standard deviations of the delays between
keystrokes and mouse button presses, the average and standard deviations of the
number of total events contained in a digraph and a trigraph are calculated.
These features are proposed and explained by Epp, Lippold y Mandryk \cite{epp2011identifying}. Most
of the times, a digraph should contain four events, while a trigraph six.
However, sometimes an individual can start a digraph or a trigraph before ending
the previous one. These additional features represent these particular cases,
and could be meaningful for the estimation of a learner's affective states.
Regarding the mouse movements, the average and standard deviation of the
duration of each mouse movement, and the averages and standard deviations of the
movements in the axes X and Y are also calculated. Lastly, a final feature is
added to preprocessing of the data. The web tutorial recorded  and included in the feature vector how many attempts
a student required before successfully solving an exercise. The final feature vector
consists of 39 features; these are based on the work of Epp, Lippold \& Mandryk
\cite{epp2011identifying} and are shown in Table \ref{tab_features}.

Once feature vectors are obtained from an experiment, the generated dataset is
normally used as training data for a classifier. Researchers of affective
recognition have used a wide variety of classification algorithms. As a proof of
concept four well known classification algorithms where compared: k-Nearest
Neighbors (k-NN) algorithm, a feed forward neural network trained with back-
propagation, a na\"ive Bayes classifier and finally a decision trees algorithm for
rational data (J-48) \cite{tan2006introduction}. Experiments and results will be presented next.

\section{Experiment and results}
\label{sec:exp}
The aim of this research is to evaluate the use of keyboard
and mouse dynamics as an appropriate sensory input for an affective recognition
system. The context of use is a learning environment for programmers. In
particular for learning activities consisting in writing short programs
interactively. An experimental approach was adopted with this aim: Sensory and
quantitative data was collected from learners as they were enrolled in a basic
course of Python programming. This data was then pre-processed using the method
described earlier in Section \ref{sec:method}, and together with the quantitative data obtained from users,
classifiers were trained and validated. The results were then compared
and we arrived to some conclusions regarding the effectiveness of the method. The goal given to
subjects was to solve as many exercises as they could in a period of two weeks.
There was neither time limit nor a minimum amount of time required for a
participant while trying to solve the exercises or complete the tutorial. The
participants were able to stop and resume their interaction with the system at
any time.
%
\begin{figure}[!t] 
\centering 
\includegraphics[width=2in]{Completed.png} 
\includegraphics[width=2in]{classDist.png} 
\caption{Completed activities and exercises and by learner in ascending order (left) and lass distribution of emotions as reported by learners (right).}
\label{fig_completed} 
\end{figure}

@@

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network, but the fact that they needed Facebook kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services; however, the use of Facebook gave researchers access to public
information about the subjects. Their ages were in the range of 18 to
30 years. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, participants had
different levels of experience, from 0 (8 persons) to more than 5 (22).

% \begin{table*}[!t]
% \centering
% \caption{ Survey question: How many years of programming experience do you have? }
% \label{tab_results}
%     \begin{tabular}{ | c | l | }
%     \hline
%     Years Programming          & Number of Learners \\
%     \hline
%    		0   &  8 \\
%     \hline
%     	1   &  13\\
%     \hline
%     	2-4  & 15\\
%     \hline
%     	5+   & 22\\
%     \hline
%     \end{tabular}

% \end{table*}

\begin{table*}[!t]
\centering
\caption{Performance of the each of the classifiers with 10-fold cross validation. Accuracy ($\kappa$) }
\label{tab_performance}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Na\"ive Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% 0.50 & 80.48\% 0.51 & 76.14\% 0.43 & 78.43\% 0.45 \\
    \hline
    Relaxation      & 71.10\% 0.40 & 72.57\% 0.45 & 69.14\% 0.37 & 71.24\% 0.34\\
    \hline
    Distraction     & 70.33\% 0.43 & 71.00\% 0.43 & 74.00\% 0.50 & 72.52\% 0.39\\
    \hline
    Frustration     & 74.71\% 0.49 & 73.86\% 0.45 & 72.62\% 0.42 & 78.00\% 0.50\\
    \hline
    Boredom         & 74.71\% 0.47 & 84.57\% 0.59 & 83.81\% 0.58 & 76.19\% 0.45\\
    \hline
    \end{tabular}
\end{table*}

@@

Table \ref{tab_performance} shows the performance of each of the four classifiers is together 
with the $\kappa$ coefficients. A 10-Fold cross validation was used. 
The accuracies and $\kappa$ coefficients obtained are close to what is
usually obtained in fixed-text methods, for example, the results presented in
Epp, Lippold y Mandryk \cite{epp2011identifying}. %A reference!!! - JJ  
												 %Added	
In a fixed-text method, subjects are asked to write specific texts.											 
The results obtained with this method are satisfactory, considering that learners
were writing a program, a task comparable with free-text writing, where there
is no restriction on what has to be written. % don't get this. What free text? A program? - JJ
In the case of the $\kappa$ coefficients, it is
usual to see values below 0.2 in methods involving free-text. In this case, some
values of $\kappa$ were close to 0.5, is important to consider the values of $\kappa$
in these results because the class distribution is not uniformly distributed. As
it is observed in many works in affective recognition, decision trees normally
produce competitive accuracy. In this case the J-48 classifier obtained an
accuracy of 80.48\%, which is marginally better than the others, and an accepted $\kappa$
statistic for the kind of problem. The artificial neural network % Which one? - JJ
give the highest
accuracy in the classification of frustration.

\section{Conclusion}
The aim of this research was to evaluate the use of keyboard and mouse dynamics 
as an appropriate sensory input for an affective recognition system. The context
of use would be learning environment for programmers. In particular for learning
activities consisting in writing short programs interactively. An experimental
approach was adopted with this aim. The proposed method in this work obtained
satisfactory results. It is usual for a classification method based on free-text
to obtain an accuracy and $\kappa$ below their counterparts of fixed-text, and in this
work the results obtained were similar to those works applied to fixed-text 
dynamics. A possible explanation of this would be the addition of the mouse 
dynamics features and the additional preprocessing performed on the feature vectors. 
While these are promising results further experiments are needed. Other experiments
should focus on novice programmers.
% But you have to comment on how this addresses the problem. Which
% should be in the title that is not there yet - JJ
% You also have to draw a _hard_ conclusion. Is programming using free
% text better than what? Typing a fixed text? 
% Is 80% effective?
